Citation Analysis of Computer Science Publications

### 2019-12-06 Scientometrics/comp.tex
fixed minor punctuation

\subsection{Data}
 A stable release of the DBLP computer science bibliography~\cite{dblp_ref} consisting of 7,079,994 records was downloaded as dblp-2018-08-01.xml.gz. Publications were parsed from the xml source file and loaded into a PostgreSQL database. We have previously parsed, the Scopus dataset of over 72,000,000 publications into a custom schema as part of a larger data platform for research evaluation~\cite{GithubERNIE2019}. Records in the DBLP dataset, were matched to Scopus identifiers using digital object identifiers (DOIs). This procedure resulted in a dataset of 2,685,356 DBLP publications with Scopus identifiers where 1,278,322 (47.6\%) were labeled as article and 1,407,034 (52.4\%) as conference proceedings. References cited by these publications were then extracted from Scopus (7,129,006), resulting in a total of 8,000,411 nodes (publications and references) and  44,296,381 edges that represent citations within the dataset. This dataset is referred to as \emph{comp} (see Fig~\ref{fig:ar_cp_annotation}).

@@

\paragraph{Clustering by Direct Citation} Graclus is a spectral graph clustering package that computes normalized cut and ratio association for a given undirected graph~\cite{graclus_2007}.  We used v1.2 in our experiments. The \emph{comp} dataset was formatted as an undirected graph, stored in a file with a header line indicating the number of nodes and edges, and used as input to Graclus, which requires the number of clusters to be formed as an input parameter. In preliminary experiments, we varied the number of clusters to be formed between 10 and 50 clusters (data not shown).  At 20 clusters, the most evenly sized set of clusters resulted with the largest cluster containing roughly 10 times the number of nodes in the smallest one~\cite{traag_louvain_2019}.  Consequently, we generated 20 clusters, labeled 0-19, using Graclus (Fig~\ref{fig:heatmap}). The quality of clustering was evaluated by measuring conductance as defined in Shun et al.~\cite{shun_parallel_2016}. Data in \emph{comp} were clustered into 18, 20, or 22 clusters and in each case the highest numbered cluster (17, 19, or 21) had the greatest conductance value and the smallest number of nodes suggesting that it effectively served as a container for `left over publications' during the Graclus process. Thus, we did not consider cluster 19 (the 20th cluster) when interpreting our results.  The remaining 19 clusters had conductance values ranging from 0.09 to 0.25 with a median conductance of 0.15.

@@

\paragraph{Clustering by Co-Citation} To build clusters using co-citations, we first implemented variable level clustering, an approach developed by Small and Sweeney (1985)~\cite{small_clustering_1985}, with minor modifications. Variable level clustering involves applying a threshold (below which all edges in a graph are dissolved) then iteratively selecting edges with the highest normalized co-citation value and extracting connected components from the graph as clusters for each edge in turn. Three parameters are needed (i) a threshold or starting level based on a quantile of normalized co-citation frequency (ii) a level increment (iii) a maximum cluster size. An issue is the generation of very large clusters by chaining through low edge weights.  Thus, at each iteration, any cluster exceeding the maximum cluster size returned to the process and a higher threshold applied to break such clusters.

In our implementation of variable level clustering, we first calculated the number of citations accumulated across Scopus for all 2,685,356 articles in \emph{comp}. After discarding those publications without any citations, \textcolor{blue}{Sitaram please confirm}, we restricted further analysis to those in the 90th percentile of citations or higher resulting in a dataset of size 212,311. We then identified 4,318,305 publications in Scopus that cite these 212,311 articles from \emph{comp}. For each of the 4,318,305 citing publications in turn, all possible reference pairs, ${n \choose 2}$, were generated where $n$ is the number of references in a publication. The cited reference pairs were restricted to those in the set of 212,311 highly cited papers previously identified.The frequency of these co-cited pairs was then computed across the dataset  and normalized using Salton's cosine formula~\cite{salton_citation_1979}. A total of 46,463,117 unique co-cited pairs were thus obtained. These data were represented in a graph where each node was a publication and the weighted edge between the pair was the normalized co-citation frequency. \par

We applied variable clustering with initial parameters of threshold ($t$) = median normalized co-citation frequency (quantile=0.5), increment $i = 0.1$, and maximum cluster size, $mcs=200$. Thus, at start, all edges below the median normalized co-citation frequency are dissolved. Clusters were formed by assembling connected components from each co-cited pair beginning with the heaviest edge weight. Clusters below size 100 were retained and any cluster larger than 200 nodes was carried over to the next round. The threshold, $t$, was then incremented by 0.1 and the process repeated while progressively incrementing $t$.  We used a bi-phasic approach where in which $t$ ranged from 0.5-0.9, after which $i$ was reduced to 0.01 for the range $0.9 \leq t \leq 0.99$. A final threshold of $t$=0.999 was applied to break the single remaining large cluster. All clusters generated thus, contained  less than 100 nodes.  Using this approach, 22,232 clusters containing 84,591 nodes were generated. Clusters containing only 2 nodes were then discarded bringing the total number of clusters down to 10,298.

Agglomerative clustering was then performed on these 10,298 clusters to generate higher-order clusters. Briefly, each cluster was now treated as a node and the edge weight between two clusters was assigned as the maximum edge weight of all edges between the nodes in these two clusters. Edges were arranged in descending order. The first pair of clusters were merged and its edge weight with other interacting clusters was recalculated. All edges were then re-ordered as before and the next pair of clusters was merged. The process was halted after 600 rounds to prevent single large clusters being generated (figure).

>>>>

\subsection{Data}
 A stable release of the DBLP computer science bibliography~\cite{dblp_ref} consisting of 7,079,994 records was downloaded as dblp-2018-08-01.xml.gz. Publications were parsed from the xml source file and loaded into a PostgreSQL database. We have previously parsed the Scopus dataset of over 72,000,000 publications into a custom schema as part of a larger data platform for research evaluation~\cite{GithubERNIE2019}. Records in the DBLP dataset were matched to Scopus identifiers using digital object identifiers (DOIs). This procedure resulted in a dataset of 2,685,356 DBLP publications with Scopus identifiers where 1,278,322 (47.6\%) were labeled as article and 1,407,034 (52.4\%) as conference proceedings. References cited by these publications were then extracted from Scopus (7,129,006), resulting in a total of 8,000,411 nodes (publications and references) and  44,296,381 edges that represent citations within the dataset. This dataset is referred to as \emph{comp} (see Fig~\ref{fig:ar_cp_annotation}).

@@

\paragraph{Clustering by Direct Citation} Graclus is a spectral graph clustering package that computes normalized cut and ratio association for a given undirected graph~\cite{graclus_2007}.  We used v1.2 in our experiments. The \emph{comp} dataset was formatted as an undirected graph, stored in a file with a header line indicating the number of nodes and edges, and used as input to Graclus, which requires the number of clusters to be formed as an input parameter. In preliminary experiments, we varied the number of clusters to be formed between 10 and 50 clusters (data not shown).  At 20 clusters, the most evenly sized set of clusters resulted with the largest cluster containing roughly 10 times the number of nodes in the smallest one~\cite{traag_louvain_2019}.  Consequently, we generated 20 clusters, labeled 0-19, using Graclus (Fig~\ref{fig:heatmap}). The quality of clustering was evaluated by measuring conductance as defined in Shun et al.~\cite{shun_parallel_2016}. 
%George, Tandy - add text about why conductance is good, and that small values are better
Data in \emph{comp} were clustered into 18, 20, or 22 clusters and in each case the highest numbered cluster (17, 19, or 21) had the greatest conductance value and the smallest number of nodes, suggesting that it effectively served as a container for `left over publications' during the Graclus process. Thus, we did not consider cluster 19 (the 20th cluster) when interpreting our results.  The remaining 19 clusters had conductance values ranging from 0.09 to 0.25 with a median conductance of 0.15.

@@

\paragraph{Clustering by Co-Citation} To build clusters using co-citations, we first implemented variable level clustering, an approach developed by Small and Sweeney (1985)~\cite{small_clustering_1985}, with minor modifications. Variable level clustering involves applying a threshold (below which all edges in a graph are deleted) then iteratively selecting edges with the highest normalized co-citation value and extracting connected components from the graph as clusters for each edge in turn. Three parameters are needed: (i) a threshold or starting level based on a quantile of normalized co-citation frequency, (ii) a level increment, and (iii) a maximum cluster size. An issue is the generation of very large clusters by chaining through low edge weights.  Thus, at each iteration, any cluster exceeding the maximum cluster size  is returned to the process and a higher threshold is applied to break such clusters.

In our implementation of variable level clustering, we first calculated the number of citations accumulated across Scopus for all 2,685,356 articles in \emph{comp}. After discarding those publications without any citations, \textcolor{blue}{Sitaram please confirm}, we restricted further analysis to those in the 90th percentile of citations or higher, resulting in a dataset of size 212,311. We then identified 4,318,305 publications in Scopus that cite these 212,311 articles from \emph{comp}. For each of the 4,318,305 citing publications in turn, all possible ${n \choose 2}$ reference pairs were generated, where $n$ is the number of references in a publication. The cited reference pairs were restricted to those in the set of 212,311 highly cited papers previously identified. The frequency of these co-cited pairs was then computed across the dataset  and normalized using Salton's cosine formula~\cite{salton_citation_1979}. A total of 46,463,117 unique co-cited pairs were thus obtained. These data were represented in a graph where each node was a publication and the weighted edge between the pair was the normalized co-citation frequency. \par

We applied variable clustering with initial parameters of threshold ($t$) = median normalized co-citation frequency (quantile=0.5), increment $i = 0.1$, and maximum cluster size, $mcs=200$. Thus, at start, all edges below the median normalized co-citation frequency were deleted. Clusters were formed by assembling connected components from each co-cited pair beginning with the heaviest edge weight. Clusters below size 100 were retained and any cluster larger than 200 nodes was carried over to the next round. The threshold, $t$, was then incremented by 0.1 and the process repeated while progressively incrementing $t$.  We used a bi-phasic approach where in which $t$ ranged from 0.5-0.9, after which $i$ was reduced to 0.01 for the range $0.9 \leq t \leq 0.99$. A final threshold of $t$=0.999 was applied to break the single remaining large cluster. All clusters generated thus, contained  less than 100 nodes.  Using this approach, 22,232 clusters containing 84,591 nodes were generated. Clusters containing only 2 nodes were then discarded bringing the total number of clusters down to 10,298.

Agglomerative clustering was then performed on these 10,298 clusters to generate higher-order clusters. Briefly, each cluster was now treated as a node and the edge weight between two clusters was assigned to the maximum edge weight of all edges between the nodes in these two clusters. Edges were arranged in descending order. The first pair of clusters were merged and its edge weight with other interacting clusters was recalculated. All edges were then re-ordered as before and the next pair of clusters was merged. The process was halted after 600 rounds to prevent single large clusters being generated (figure).
